Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
typo
minor typos
fixed all the reviews from Dan

\begin{figure*}[ht!]
		\centering
		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
		\caption{Runtime Memory and Simulated Counter-Part\vspace{-0.2cm}}
		\label{fig:SimulatedMemory}
\end{figure*}

@@

On higher addresses, we can find the beginning of the memory managed by the VM. At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.

@@

The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code. When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when eden is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine. The runtime can then access the current machine code stack frame though these variables. In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text. Referencing these variables and invoking runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible. First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

@@

To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulation byte array.  This includes mock addresses for all the object the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the in-image compilation process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.

@@

Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecode method. Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.

@@

Recently, to evaluate a new old space garbage collection algorithms, we decided to implement standard algorithms to get reference benchmark results. Since we had already a Mark-Compact collector, we mainly introduced a Mark-Sweep non moving GC as an alternative. The whole implementation was done in the simulator, and only when it was working there, was it compiled to C. Using this process, the compiled C code was working out of the box. %To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

We started by sketching the new Sweep algorithm in the code base. Once it was partially written, we started the simulator. Since the algorithm was not complete, we added the missing pieces inside the debugger, installing the new code at runtime and checking against the current values of the function parameters if the function behaved as expected, as one can do it in any Smalltalk application.

In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory, performs the GC there, and if no exception was raised\footnote{The VM code base includes many assertions to ensure specific state is correct at given points in the execution, such assertions are optionally compiled in the production VM, they represent the majority of exceptions raised during development or bug resolution.}, it then performs the GC on the original version. This means that each time the algorithm did not work correctly, the erratic behavior happened in a copy of the heap. Since we can clone again the original version as many times as we want, we were able, each time the memory got corrupted beyond repair, to restart the exact same GC execution on the exact same heap with each objects at the exact same addresses. Indeed, in most cases when the bug happens, the memory was already very corrupted and it was very difficult to track back where the bug came from without rewinding the memory state. We reproduced the exact same GC many times in different cases to fully understand specific bugs.

@@

The debugging process discussed here is exampled in a youtube video\footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.

@@

\subsection{Machine code zone send analysis}

@@

One of the first analysis we ran was on the machine code zone. We stopped the simulation when the machine code zone reached 1Mb. We then iterated over it and investigated what was in. As show in Table \ref{tab:generalAnalysis}, 1752 methods were compiled to machine code by the JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (basically, they have never been used).

@@

Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.

@@

The code used for these analysis is detailed in the Section "Let Me Tell You All About It, Let Me Quantify" of the blog post "Build me a JIT as fast as you can"\footnote{http://www.mirandabanda.org/cogblog/2011/03/01/build-me-a-jit-as-fast-as-you-can/}.

@@

 The simulator has several limitations. Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production. However, we have been using this infrastructure since 1995, and these kinds of bugs are really rare and usually easy to fix. We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot (yet) be simulated.

\paragraph{Performance.}
The first limitation is due to the simulation performance. The interpreter-only is simulator is between a hundred and a thousand times slower to execute code than the normal VM. With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation performance usually drops by a factor of two, though in specific cases it is actually faster. We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.

In practice, if a GC bug happens in an application 15 minutes after start-up, it takes around 50 hours to reproduce in the simulator. Fortunately, we work around this problem by using snapshots. Once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes. The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds. In addition, since the interpreter-only simulator is usually quicker at executing code, if the bug is unrelated to the JIT (typically, a GC bug), the interpreter-only simulator can be used.

\paragraph{Calls to external code.}
Although most of the GC and JIT development and debugging can be done in the simulator, specific tasks cannot be done this way. Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated. For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation. However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).

The main limitation we have right now is with Foreign Function Interfaces (FFI). We have a significant amount of bugs in FFI, often due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects. Such bugs cannot be debugged with our simulation infrastructure so far and we have to rely on gdb/lldb.

@@

Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. For example, the teams working on the Truffle/Graal projects \cite{OneVMToRule} have very convenient development tools for their VMs such as a tool to visualize the intermediate representation of their JIT Graal at different point in the compilation process. But, despite an impressive number of publications on the projects, including publications on development tools for the guest languages \cite{TrufflePolyglotDebugger} or on specific parts of the VM \cite{GraalIR2}, no publication seems to be directly related to the VM development tools.

@@

\paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18. They allow to inspect the running state of the Maxine VM while it runs for debugging purposes. One of the main difference with our design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode. We believe having two different modes allows us to easily generate a production VM while still having nice debugging features. Having a full metacircular VM would be interesting, but it is unclear whether it is convenient to build a production VM in that way. So far, most production VMs (Java, Javascript, etc.) are still compiling through the C/C++ compiler and are not metacircular.

@@

In the near future, we plan to extend the simulator with customizable development tools. Currently the tools built in top of the simulation infrastructure are mainly textual. More advanced tooling such as the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} should enable a more interactive interface, making it easier to apprehend by new developers.

>>>>

\begin{figure*}[ht!]
		\centering
		\includegraphics[width=0.88\linewidth]{figures/SimulatedMemory}
		\caption{Runtime Memory and Simulated Counter-Part\vspace{-0.2cm}}
		\label{fig:SimulatedMemory}
\end{figure*}

@@

On higher addresses, we can find the beginning of the memory managed by the VM. At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (this optional execution exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.

@@

The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code. When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when young space is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine. The runtime can then access the current machine code stack frame though these variables. In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text. Referencing these variables and invoking runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible. First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

@@

To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulation byte array.  This includes mock addresses for all the objects the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the in-image compilation process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code; adaptive optimizations and speculative optimizations are debugged differently.

@@

Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecode method. This is hardly pleasurable for methods of the complexity produced by the adaptive optimiser though.

@@

Recently, to evaluate some new old space garbage collection algorithms, we decided to implement standard algorithms to get reference benchmark results. Since we had already a Mark-Compact collector, we mainly introduced a Mark-Sweep non moving GC as an alternative. The whole implementation was done in the simulator, and only when it was working there, was it compiled to C. Using this process, the compiled C code was working out of the box. %To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

We started by sketching the new Sweep algorithm in the code base. Once it was partially written, we started the simulator. Since the algorithm was not complete, we added the missing pieces inside the debugger, installing the new code at runtime and checking against the current values of the function parameters if the function behaved as expected, as one can do in any Smalltalk application.

In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory, performs the GC there, and if no exception was raised\footnote{The VM code base includes many assertions to ensure specific state is correct at given points in the execution, such assertions are optionally compiled in the production VM, they represent the majority of exceptions raised during development or bug resolution.}, it then performs the GC on the original version. This means that each time the algorithm did not work correctly, the erratic behavior happened in a copy of the heap. Since we can clone again the original version as many times as we want, we were able, each time the memory got corrupted beyond repair, to restart the exact same GC execution on the exact same heap with each object at the exact same addresses. Indeed, in most cases when a bug happens, the memory was already corrupted and it was very difficult to track back where the bug came from without rewinding the memory state. We reproduced the exact same GC many times in different cases to fully understand specific bugs.

@@

The debugging process discussed here is illustrated in a youtube video\footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.

@@

\subsection{Analyzing the machine code zone}

@@

One of the first analyses we ran was on the machine code zone. We stopped the simulation when the machine code zone reached 1Mb. We then iterated over it and investigated what was in it. As show in Table \ref{tab:generalAnalysis}, 1752 methods were compiled to machine code by the JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (basically, they have never been used).

@@

Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle's statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.

@@

The code used for these analyses is detailed in the Section "Let Me Tell You All About It, Let Me Quantify" of the blog post "Build me a JIT as fast as you can"\footnote{http://www.mirandabanda.org/cogblog/2011/03/01/build-me-a-jit-as-fast-as-you-can/}.

@@

 The simulator has several limitations. Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production. However, we have been using this infrastructure since 1995, and these kinds of bugs are rare and usually easy to fix. We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot (yet) be simulated.

\paragraph{Performance.}
The first limitation is due to the simulation performance. The interpreter-only simulator is between a hundred and a thousand times slower to execute code than the normal VM. With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation performance usually drops by a factor of two, though in specific cases it is actually faster. We usually enable conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.

In practice, if a GC bug happens in an application 15 minutes after start-up, it takes around 50 hours to reproduce in the simulator. Fortunately, we work around this problem by using snapshots. Once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes. The VM simulator can then be started just before the crash and the debugging tools can be used in far less than a minute. In addition, since the interpreter-only simulator is usually quicker at executing code, if the bug is unrelated to the JIT (typically, a GC bug), the interpreter-only simulator can be used.

\paragraph{Calls to external code.}
Although most of the GC and JIT development and debugging can be done in the simulator, some specific tasks cannot be done this way. Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated. For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation. However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator; that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).

We have a significant amount of bugs in Foreign Function Interfaces (FFI), often due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects. Such bugs cannot be debugged with our simulation infrastructure so far and we have to rely on gdb/lldb.

@@

Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. For example, the teams working on the Truffle/Graal projects \cite{OneVMToRule} have very convenient development tools for their VMs such as a tool to visualize the intermediate representation of their JIT Graal at different points in the compilation process. But, despite an impressive number of publications on the projects, including publications on development tools for the guest languages \cite{TrufflePolyglotDebugger} or on specific parts of the VM \cite{GraalIR2}, no publication seems to be directly related to the VM development tools.

@@

\paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18. They allow one to inspect the running state of the Maxine VM while it runs for debugging purposes. One of the main differences with our design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode. We believe having two different modes allows us to easily generate a production VM while still having nice debugging features. Having a full metacircular VM would be interesting, but it is unclear whether it is convenient to build a production VM in that way. So far, most production VMs (Java, Javascript, etc.) are still compiling through the C/C++ compiler and are not metacircular.

@@

In the near future, we plan to extend the simulator with customizable development tools. Currently the tools built on top of the simulation infrastructure are mainly textual. More advanced tooling such as moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} should enable a more interactive interface, making it easier to apprehend by new developers.
